\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-29 coqimpl.tex
Reply comments first so that yizhou can continue to read

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$
  \begin{itemize}
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,  @\YZ{Can R be indexed?}\EDJreply{Yes. Discussed below. Basically constructing left inverse for simple inductive type, this weaker form is enough}@
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), .., 
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of
  "C" and "prec" of "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}

@@

We can support partial dependent eliminator as well, but partial dependent eliminator can deduce "prec" and our "prec" is enough to contruct this left inverse for (non-indexed) inductive type. 
\begin{minted}{Coq}
  pdelim : ∀ (P : D -> Set), .., 
  (rᵢ : (∀ (x : Aᵢ) (w : Bᵢ x -> D), (∀ (b : Bᵢ x), P (w b)) -> (P (dᵢ x w)))),  
  .., ∀ (d : D) -> option (P d)
\end{minted}
To generalize to indexed inductive type, we base on indexed W type~\cite{martin1982constructive, morris2009indexed,jashug2017}. Please refer to

>>>>

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$
  \begin{itemize}
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,  @\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}@
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), .., 
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of
  "C" and "prec" of "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}

@@

We can support partial dependent eliminator as well, but partial dependent eliminator can deduce "prec" and our "prec" is enough to contruct this left inverse for (non-indexed) inductive type. 
\begin{minted}{Coq}
  pdelim : ∀ (P : D -> Set), .., 
  (rᵢ : (∀ (x : Aᵢ) (w : Bᵢ x -> D), (∀ (b : Bᵢ x), option (P (w b))) -> option (P (dᵢ x w)))),  
  .., ∀ (d : D) -> option (P d)
\end{minted}
To generalize to indexed inductive type, we base on indexed W type~\cite{martin1982constructive, morris2009indexed,jashug2017}. Please refer to

### 2022-06-29 metatheory.tex
comments
Reply comments first so that yizhou can continue to read

We start with the formulation of the syntax, which is based on the formalization of a predicative \textit{Martin-Lof Type Theory} (MLTT) given by~\citet{coquand2018canonicity}. Thus we will omit a great deal of details about MLTT here but focus on the new facility we introduced.

@@

In our syntax, each judgement will actually be represented by a QIIT type, which are
mutually recursively defined together. Quotient is used to represent
judgemental(definitional) equality, thanks to which we can have a
concise representation. This way, for all four kinds of judgement we are
using, there are always equality among these data. Doing so, we can omit
some obvious equality judgement. Most of the reasoning can be formulated
using the \textit{algebra} of QIIT (mapping out function from this
quotient data-type).

We also highlight some special feature in this style of formulation of dependent type theory: 1. this modern formulation of type theory does't have operational semantic but only equality 
%We will recover canonicity afterwards, which implies the computational ability of our theory. 
;
2. instead of using meta-level substitution, we use explicit substitution and de Bruijn indices in our intrinsic formulation. This is also called substitution calculus in the literature and is favoured due to its clear categorical semantic; 
3. we are dealing with quotient data, so we are actually manipulating equivalence classes of terms.

We will write out the syntax of the whole type theory for completeness.

>>>>

% Motivation of using QIIT
We start with the formulation of the syntax, which is based on the
formalization of a predicative \textit{Martin-Lof Type Theory} (MLTT)
given by~\citet{coquand2018canonicity}. Thus we will omit a great deal
of details about MLTT here\YZ{What details are omitted?} but focus on the new facility we introduced.

@@

% In our syntax, each judgement will actually be represented by a QIIT type, which are
% mutually recursively defined together. Quotient is used to represent
% judgemental(definitional) equality, thanks to which we can have a
% concise representation. This way, for all four kinds of judgement we are
% using, there are always equality among these data. Doing so, we can omit
% some obvious equality judgement. Most of the reasoning can be formulated
% using the \textit{algebra} of QIIT (mapping out function from this
% quotient data-type).

% We also highlight some special feature in this style of formulation of dependent type theory: 1. this modern formulation of type theory does't have operational semantic but only equality 
% %We will recover canonicity afterwards, which implies the computational ability of our theory. 
% ;
% 2. instead of using meta-level substitution, we use explicit substitution and de Bruijn indices in our intrinsic formulation. This is also called substitution calculus in the literature and is favoured due to its clear categorical semantic; 
% 3. we are dealing with quotient data, so we are actually manipulating equivalence classes of terms.

% We will write out the syntax of the whole type theory for completeness.

### 2022-06-29 syntactic-example.tex
Reply comments first so that yizhou can continue to read

We provide one example in its surface syntax as pseudocode, and their
meta-theory encoding. This example is computing the predecessor of a
mundane natural number, but we split the inductive definition into two
parts and making "O" and "S" standing alone.
"pred" is defined as $n \mapsto ("pred" n, n)$\YZ{looks like an ill-founded definition}, and thus a predecessor.

@@

 EndFamily. (* pN₁, handler₁, sig₃ *)
 pred@\YZ{Why is there no FRecursor?}@
  : self[A].N → self[A].N × self[A].N
  := ... 
  (* predT, pred₁, sig₄, sig₄', sl₄ *) 
 k := self[A].pred self[A].O 
(* pN₂, sig₅ *)
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.45\linewidth}
\begin{minted}[escapeinside=@@]{Coq}
Family A₂ extends A.
 Extend FInductive N := S : N → N.
 (* Nsig, sig₁₁, inh₁, O₂, sig₂₁, inh₂, 
    sig₂₂, sig₂₁', sig₂₂', sl₂₂, inh₃ *)
 Extend Family handler.
   Inherit O. (* inhᵢₙ₀, pN₃ *)
   S : self[A].N × self[A].N
       → self[A].N × self[A].N 
     := fun (x,y) => (y, self[A₂].S y).
 EndFamily. (* inhᵢₙ, sig₃₁, inh₄*)
 pred 
  : self[A₂].N → self[A₂].N×self[A₂].N 
  := ... 
  (* sig₄₁, inh₅, sl₄₁ *)
 Inherits k. 
 (* inh₆ *)
EndFamily.
    \end{minted}
  \end{minipage}
  \caption{Demonstrating Pseudo-Code}
\end{figure}

>>>>

We provide one example in its surface syntax as pseudocode, and their
meta-theory encoding. This example is computing the predecessor of a
mundane natural number, but we split the inductive definition into two
parts and making "O" and "S" standing alone.
"pred" will map $S~n \mapsto (n, S~n)$ and $O \mapsto (O, O)$\YZ{looks like an ill-founded definition}\EDJreply{Fixed}, and thus a predecessor.

@@

 EndFamily. (* pN₁, handler₁, sig₃ *)
 FRecursor pred about self[A].N 
  motive (fun _ => self[A].N × self[A].N)
  using handler.@\YZ{Why is there no FRecursor?}\EDJreply{I wanted to save space and emphasize the type. Now fixed}@
 (* pred
  : self[A].N → self[A].N × self[A].N *)
  (* predT, pred₁, sig₄, sig₄', sl₄ *) 
 k := self[A].pred self[A].O 
(* pN₂, sig₅ *)
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.45\linewidth}
\begin{minted}[escapeinside=@@]{Coq}
Family A₂ extends A.
 Extend FInductive N := S : N → N.
 (* Nsig, sig₁₁, inh₁, O₂, sig₂₁, inh₂, 
    sig₂₂, sig₂₁', sig₂₂', sl₂₂, inh₃ *)
 Extend Family handler.
   Inherit O. (* inhᵢₙ₀, pN₃ *)
   S : self[A].N × self[A].N
       → self[A].N × self[A].N 
     := fun (x,y) => (y, self[A₂].S y).
 EndFamily. (* inhᵢₙ, sig₃₁, inh₄*)
 Inherit pred.

  (* sig₄₁, inh₅, sl₄₁ *)
 Inherits k. 
 (* inh₆ *)
EndFamily.
    \end{minted}
  \end{minipage}
  \caption{Demonstrating Pseudo-Code}
\end{figure}

### 2022-06-29 syntax.tex
Reply comments first so that yizhou can continue to read

Firstly, there are four standard judgement forms in MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. Our presentation
follows the style of~\citet{kaposi2019gluing}, but for readability we
omit universe levels attached to judgements.
After these,
we will have four new kinds of judgements to model our linkages.

Because we formalize our language as a deep embedding in a type theory with QIIT types, every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms (otherwise how can they be encoded in Agda?)
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the fake Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, we spell out these \textit{well-formedness rules} of type
judgements and their Agda counterparts to make things more accessible. Of
course, a well-formed judgement is not necessarily derivable, just like
a well-typed QIIT type is not necessarily inhabited.

Due to our intrinsic setting, \textbf{we can and will omit a lot of
presumptions of the typing rules}. For example, we don't spell out that
the rule "Type Universe" requires a well-typed context, even though it
does implicitly, because it is not possible to have ill-formed
contexts in our intrinsic setting.

Among them, sealing judgement $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is the
most special one because it is actually just a shorthand of term
judgement $\goodTerm{\Gamma , \cL \sigma}{\_}{A[\pi_1]}$.\YZ{You haven't introduced the notation A[pi_1] yet at this point. And what is A[pi_1]?}

>>>>

We formalize our language as a deep embedding in a type theory with QIIT types, every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms. (Otherwise how can they be encoded in Agda?) There are four standard judgement forms in MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. These four judgements (QIIT types) are
mutually recursively defined together. Our presentation
follows the style of~\citet{kaposi2019gluing}, but for readability we
omit universe levels attached to judgements.
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the fake Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, we spell out these \textit{well-formedness rules} of type
judgements and their Agda counterparts to make things more accessible. Of
course, a well-formed judgement is not necessarily derivable, just like
a well-typed QIIT type is not necessarily inhabited.

Due to our intrinsic setting, \textbf{we can and will omit a lot of
presumptions of the typing rules}. For example, we don't spell out that
the rule \ruleref{Type Universe} requires a well-typed context, even though it
does implicitly, because it is not possible to have ill-formed
contexts in our intrinsic setting.

Quotient is used to represent
judgemental(definitional) equality (for each QIIT type), thanks to which we can have a
concise representation. This way, for all four kinds of judgement we are
using, there are always equality among these data. Doing so, we can omit
some obvious equality judgement. Most of the reasoning can be formulated
using the \textit{algebra} of QIIT (mapping out function from this
quotient data-type).

Now we highlight some other special feature contrary to more conventional setting: 1. this modern formulation of type theory does't have operational semantic but only equality\footnote{Later we will prove canonicity which can imply a certain level of computational ability of our theory}
%We will recover canonicity afterwards, which implies the computational ability of our theory. 
;
2. instead of using meta-level substitution, we use explicit substitution and de Bruijn indices in our intrinsic formulation. This is also called substitution calculus in the literature and is favoured due to its closedness categorical semantic; 
3. we are dealing with quotient data, so we are actually manipulating equivalence classes of terms.

We quickly review the substitution calculus,

Based upon the basic MLTT,
we will have four new kinds of judgements to model our linkages. Among them, sealing judgement $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is the
most special one because it is actually just a shorthand of term
judgement $\goodTerm{\Gamma , \cL \sigma}{\_}{A[\pi_1]}$.\YZ{You haven't introduced the notation A[pi_1] yet at this point. And what is A[pi_1]?}
